Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-10 Manuscript/manuscript.tex
Update introduction and electrometer concept section

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. The detailed results from simple model systems can be then
used to understand the role of ions in complex biological systems.

Zwitterionic phosphocholine (PC) lipid bilayers are commonly used model systems for
cellular membranes. Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy, give accurate information about ion binding
in PC lipid bilayers\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Interpretation most of these experiments suggests that Na$^+$ ions exhibit
negligible binding in PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules. Atomistic resolution
molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. This study presented 
a direct comparison between ion binding affinity between simulations and
experiments by using NMR data for lipid headgroup order 
parameters and the electrometer concept~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
As a starting point we use the parameters from the Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly for reproducing the membrane binding affinities and induced
structural effects of Sodium and Calcium ions.

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding between experiments and simulations was compared by 
using lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}.
The concept is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid,  $X^\pm$~\cite{seelig87}.
Change in the order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

The change of the head group order parameters is empirically quantified
as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for Ca$^{2+}$ binding to POPC
bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because recent analysis concluded 
that the order parameters decrease with bound positive charge 
and increase with bound negative charge when the signs are taken 
in account \cite{ollila16,catte16}. This is rationalized as a change
of lipid head group dipole tilt toward water phase with bound positive
charge and {\it vice versa} with negative charge \cite{seelig87}.

@@

The measured order parameter change depends on the response of 
the headgroup on the bound charge, i.e. $m_i$ in Eq.~\ref{OPchangeEQ},
and the ion binding affinity. Thus, the former property has to be
well quantified before using the electrometer concept to
analyze binding affinities. This is done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response also in simulations,
we use experimental data for dihexadecyldimethylammonium (2C$_{16}$$^+$N2C$_1$Br$^-$)
in a POPC bilayer \cite{scherer89}. Dihexadecyldimethylammonium
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems \cite{scherer87}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium \\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH2{)}_{15}-CH_3)(-[:30]{(}CH2{)}_{15}-CH_3)
} , 
\vspace{0.5cm} \\
which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
%Chloride counter ions were used
%in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
%were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.

>>>>

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
often used as model systems to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

Interactions of biological cations, especially Na$^+$ and
Ca$^{2+}$, with PC bilayers are widely studied in
experiments~\cite{akutsu81,altenbach84,seelig90,cevc90,tocanne90,binder02,pabst07,uhrikova08}
and classical MD simulations~\cite{bockmann03,bockmann04,Berkowitz12,melcrova16,javanainen17}.
The details of ion binding are, however, not fully consistent in the literature.
%not agreed in the literature \cite{cevc90,tocanne90,seelig90,binder02,catte16,??}.
%General conclusion from experimental
%studies has been that multivalent ions and lithium have weak specific binding in
%phopholipid bilayers, while other monovalent ions do not essentially
%bind 
%The presence of anoinic lipids, like PS or PG,
%increase the concentration close to the bilayer and thus the amount of bound
%ions, but do not affect the specific binding constant \cite{seelig90}.
Interpretations of non-invansive spectroscopic methods, like nuclear magnetic resonance (NMR), scattering
and infrared scpectroscopy suggest that Na$^+$ ions exhibit
negligible binding to PC lipid bilayers with submolar concentrations, while Ca$^{2+}$ specifically
binds to phosphate groups of two lipid molecules~\cite{hauser76,hauser78,herbette84,akutsu81,altenbach84,binder02,pabst07,uhrikova08}.
Atomistic resolution molecular dynamics (MD) simulation models, however, predict significantly
stronger binding for the cations than NMR experiments \cite{catte16}.
On the other hand, some experiments have also been interpreted to support the predictions from
MD simulations~\cite{bockmann03,vacha09a}. 
Furthermore, interactions of Calcium ions with 3-4 lipids,
including also interactions with carbonyl
oxygens, have been reported from simulations~\cite{bockmann03,bockmann04,melcrova16,javanainen17}.

Recent work published by the NMRlipids project (\url{nmrlipids.blogspot.fi})~\cite{catte16}
made an attempt to resolve the apparent controversies. A direct comparison of
ion binding affinities to PC bilayers was presented between simulations and
experiments by using the electrometer concept, which is based on the experimental NMR data
for the lipid headgroup order parameters~\cite{seelig87}.
Using massive amounts of data collected by Open Collaboration method, it was concluded
that the accuracy of the current state of the art lipid models for MD
simulations is not sufficient for the detailed interpretation of the cation
interactions with PC lipid bilayers~\cite{catte16}.

@@

In this work we show that the cation binding behavior
in MD simulations of 1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer can be significantly 
improved by implicitly including the electronic polarizability
in the polar region of lipid molecules. The electronic
polarizability is included by using the electronic
continuum correction (ECC)~\cite{leontyev11}, which
has been previously shown to improve the behaviour of
MD simulations of ions in bulk water~\cite{jungwirth17-new-paper-to-be-published,Pluharova2014,kohagen14,kohagen16}. 
As a starting point we use the parameters from the Lipid14 model \cite{dickson14},
which gave the best cation binding behaviour in the previous study \cite{catte16}.
The developed ECC-lipid parameters reproduce the experimentally
measurable structural parameters of an ion-free POPC lipid bilayer with the 
accuracy comparable to the other state of the art lipid models, while surpassing
them significantly for reproducing the membrane binding affinities and induced
structural effects of Sodium and Calcium ions.

@@

\subsection{Electrometer concept} \label{section:electrometer}
Ion binding in lipid bilayers was compared between experiments and simulations by 
using the lipid head group order parameters and the
''electrometer concept'' \cite{seelig87,catte16}, which
is based on the experimental observation that the C-H bond
order parameters of $\alpha$ and $\beta$ carbons in PC lipid head group
(see Fig.~\ref{simVSexpNOions}) are proportional to the amount 
of unit charge bound per lipid~\cite{seelig87}.
The change of order parameters measured with varying aqueous
ion concentration can be then related to the amount of bound ions.

@@

The relation between bound charge per lipid,  $X^\pm$, and
the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$,
is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
  \Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx \frac{4m_i }{3\chi}X^\pm,
\end{equation}
where $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge,
$i$ refers to either $\alpha$ or $\beta$ carbon,
$m_i$ is an empirical constant depending on the valency and position of the bound charge,
and the experimental value \cite{seelig77,davis83}, $\chi \approx$\,167\,kHz, is used for the quadrupole coupling constant.
Atomic absorption spectra and $^2$H NMR data gave
$m_\alpha=-20.5$~kHz  and $m_\beta=-10.0$~kHz for the binding of
Ca$^{2+}$ to POPC bilayer (in the presence of 100~mM NaCl)~\cite{altenbach84,ollila16,catte16}.
The slopes are negative, because the head group order parameters
decrease with bound positive charge and increase with bound negative
charge~\cite{ollila16,catte16}. This can be rationalized as a change
of the lipid head group dipole tilt toward water phase with the bound positive
charge and {\it vice versa} with the negative charge~\cite{seelig87}.

The measured order parameter change depends on both, the head group response 
to the bound charge and the amount of bound charge, i.e. $m_i$ and $X^\pm$
in Eq.~\ref{OPchangeEQ}, respectively. The former property has to be
well quantified before the electrometer concept can used to 
analyze the binding affinities, as done experimentally for a wide range
of systems~\cite{seelig87,beschiasvili91}. To calibrate 
the head group order parameter response to the bound charge in simulations,
we use the experimental data for dihexadecyldimethylammoniumbromide %(2C$_{16}$$^+$N2C$_1$Br$^-$)
mixed within a POPC bilayer~\cite{scherer89}. Dihexadecyldimethylammonium\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
 -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cation surfactant having two acyl chains and bearing a unit
charge in the hydrophilic end. Thus, it is expected to locate
in the bilayer similarly to the phospholipids and the molar ratio then 
gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium, which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. 
%Chloride counter ions were used
%in simulations, because parameters for bromide, which was used in the experiment \cite{scherer89},
%were not available in the standard Gromacs files for Amber force field.
The systems were simulated for 200~ns using Lipid14 model for POPC where reasonable lipid neighbor exchange occurs.
First 20~ns were omitted from the analysis.
